Reproducible Research - Peer assessment 1
========================================================

By Juan Carlos Lopez Hernandez                          Date: 15 May 2015



Introduction
============

The assigment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

I write a report that try to answer the questions detailed in this document. This report presents the results in a single **R markdown document** that can be processed by knitr and be transformed into an HTML file.


R Enviroment
============

We need to load some libraries. We use ggplot2 for plotting figures

```{r}
library(knitr)
library(data.table)
library(ggplot2) 
```

Also, When I write code chunks in the R markdown document, we always use **echo = TRUE** so everybody will be able to read the code.

```{r}
opts_chunk$set(echo = TRUE, results = 'hold')
```


Loading and preprocessing the data
==================================

The dataset is stored in a *CSV file*. This dataset contains a total of 17,568 observations.

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

**steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA).

**date**: The date on which the measurement was taken in YYYY-MM-DD format.

**interval**: Identifier for the 5-minute interval in which measurement was taken.

**NOTE**: The GitHub repository also contains the dataset for the assignment so you do not have to download the data separately. If you want to download the data yo can put the file in the current working directory

Setting working directory and reading data file

```{r}
setwd("C:/Users/user/Documents/R/5 Reproducible Research/Data_Activity")

data_activity <- read.csv('activity.csv', header = TRUE, sep = ",",
                          colClasses=c("numeric", "character", "numeric")
                         )
```

We convert the date field into to Date class. 
Also we convert the Interval field to Factor class.

```{r}
data_activity$date <- as.Date(data_activity$date, format = "%Y-%m-%d")
data_activity$interval <- as.factor(data_activity$interval)
```

We check data using str():

```{r}
str(data_activity)
```

and summary() methods

```{r}
summary(data_activity)
```


What is mean total number of steps taken per day?
=================================================

For answer this question we ignore the missing values.
First, We must calculate the total steps per day.

```{r}
number_steps_day <- aggregate(steps ~ date, data_activity, sum)
colnames(number_steps_day) <- c("date","steps")
head(number_steps_day)
```

Now we calculate the mean and median of the number of steps taken per day.
We exclude missing values from analyses of mean & median:

```{r}
steps_mean   <- mean(number_steps_day$steps, na.rm=TRUE)
steps_median <- median(number_steps_day$steps, na.rm=TRUE)
paste("The mean total number of steps taken per day is:",steps_mean)
paste("And the median total number of steps taken per day is:",steps_median)
```

We make at this moment an histogram of the total number of steps taken per day, plotted with appropriate bin interval.

```{r}
ggplot(number_steps_day, aes(x = steps)) + 
       geom_histogram(fill = "blue", binwidth = 1100) + 
        labs(title="Histogram: Steps taken per day", 
             x = "Number of Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```


What is the average daily activity pattern?
===========================================

1. We make the plot with the time series of the average number of steps taken, averaged across all days (y-axis):

```{r}
steps_per_interval <- aggregate(data_activity$steps, 
                                by = list(interval = data_activity$interval),
                                FUN=mean, na.rm=TRUE)

#We convert values to integers for helping in plotting

steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```

```{r}
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="blue", size=1) +  
        labs(title="Average daily activity", x="Interval", y="Number of steps") +  
        theme_bw()
```


2. We need to answer: Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

First, we calculate the 5-minute interval with the containing the maximum number of steps: 
```{r}
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]

paste("The "," has maximum max_interval ",max_interval,"steps.")
```
